Software Citation Principles

### 2016-03-24 sc-principles/software-citation-principles.tex
added my middle initial
More sub-sections

These software citation principles were originally based on an adaptation of the FORCE11 Data Citation Principles~\cite{data-citation-principles}, and then were modified based on working group discussions and the information from the use cases and the related work.
The adaptations and modifications have been made because software, while similar to data in terms of not traditionally having been cited in publications, is also different than data in that it can be used to express or explain concepts, it is updated more frequently, and it is executable.  \katznote{are there any other difference between software and data that should be stated here?}  Also, while software can be considered a type of data, the converse is not generally true.

@@

With close to 50 working group participants (see Appendix~\ref{app:wg_members}) representing a range of research domains, the working group was tasked to
document existing practices in their respective communities. A total of 47 documents were submitted by working group
participants, with the life sciences, astrophysics, and geosciences being particularly well-represented in the submitted
resources.

@@

Some of the most actionable work has come from the UK Software Sustainability Institute (SSI) in the form of blog posts written by their community fellows:

@@

Wilson suggests that software authors include a `CITATION' file that documents exactly how the
authors of the software would like to be cited by others~\cite{ssi-citation-files}. While this isn't a formal metadata specification (it's not machine
readable for example) this does offer a solution for authors wishing to give explicit instructions to potential citing authors and as noted in the introduction (\S\ref{sec:intro}), there is evidence that authors follow instructions if they exist~\cite{10.1371/journal.pone.0136631}.

@@

Most notable of the community efforts are those of WSSSPE~\cite{wssspe} and
SSI~\cite{ssi-workshops} who between them have run a series of workshops aimed
at gathering together community members with an interest in improving the
status of software in academia. \asnote{Not sure this is a sufficient one-line
summary but couldn't think of anything better.} In each of the three years that
WSSSPE workshops have run thus far, the participants have produced a report
documenting the topics covered. Section 5.8 and Appendix J in \cite{WSSSPE3}
has some preliminary work and discussion particularly relevant to this working
group.

@@

One approach to increasing the `citability' of software is to encourage the
submission of papers in standard journals describing a piece of research
software, often known as software papers (see \S\ref{sec:software_papers}).
While some journals (e.g., Transactions on Mathematical Software [TOMS], Bioinformatics, F1000Research) have traditionally accepted
software submissions, the American Astronomical Society (AAS) have recently
announced they will accept software papers in their journals~\cite{aas-sofware-papers}. 
Professional societies are in a good position to change their respective communities, as the publishers of journals and conveners of domain-specific conferences; as publishers they can change editorial policies (as AAS has done) and conferences are an opportunity to communicate and discuss these changes with their communities.

In astronomy \& astrophysics: ASCL~\cite{ascl}, AAS journals  and the joint
AAS/GitHub workshop~\cite{aas-software-index} dedicated to software citation,
indexing and discoverability, and the report from a Birds of a Feather session
organized by the ASCL at the Astronomical Data Analysis Software and Systems
(ADASS) XXV conference~\cite{2015arXiv151207919A}.

Life sciences/NIH: Software Discovery Index Meeting Report~\cite{software-discovery-index} a cross-disciplinary meeting
but with a largely biomedical-focus. Suggested minimal information about software (MIAS) standard. Discusses how a common set of metadata
fields are critical for useful indexing - suggested candidates for this list.

Geosciences: Ontosoft~\cite{ontosoft} - A Community Software Commons for the Geosciences. Lots of attention given to the metadata
required to describe, discover and execute research software. NSF geosciences workshop around data lifecycle, management and citation~\cite{nsf-geo-data}. This report includes many recommendations for data citation.
\LJHnote{ The Seismological Research Letters, Electronic Seismologists publishes papers on software.} %text needs to be cleaned up LJH%

@@

Van de Sompel et al.~\cite{VandeSompel2004} argued for registration of all units of scholarly communication, including software.\niemnote{suggesting this reference go here}
Author guidelines in F1000Research, APA authors, SSI `publish or be damned'
\cite{ssi-publish-or-be-damned} manifesto, workflow for publishing code using
GitHub and Zenodo \cite{github-citable-code-guide}.
\dmnote{There are many other journals publishing software descriptions, some even specialized in software, e.g. \href{http://www.openresearchcomputation.com}{Open Research Computation}.}

\textbf{Existing efforts around metadata standards}

@@

\textbf{Studies of author motivations}

@@

\subsection{What software to cite}

The software citation principles do not define what software should be cited, but rather, how software should be cited.
What software should be cited is the decision of the author(s) of the research work, in the context of community norms and practices, and in most scientific communities, these are currently in flux.
In general, we believe that any software on which a research product directly depends should be cited, if it significantly contributes to the research product.
Again, the specific decision of what is significant must be made by the author(s) of the product.
However, an illustrative example is the use of Microsoft Excel in research.
We suggest that if Excel is used to simply store and plot data, it does not need to be cited, but if it is used for statistical analysis, it should be.  Similarly, general software for conducting library research (e.g., JSTOR Mobile App), writing research papers (e.g., \LaTeX), scientific presentations (e.g., Powerpoint) or communications (e.g., Skype) should not be cited.
This recommendation matches that of the Purdue Online Writing Lab: ``Do not cite standard office software (e.g. Word, Excel) or programming languages.  Provide references only for specialized software.''~\cite{powl-citing-software}

Note that some software which is or could be captured as part of data provenance may not be cited.
Citation is a record of software that is important to the research outcome, where provenance is a record of all steps (including software) used to generated particular data within the research process.
This implies that for a data research product, provenance data will include all cited software, but not necessarily vice versa.
Similarly, the software metadata that is recorded as part of data provenance should be a superset of the metadata recorded as part of software citation.
And the data recorded for reproducibility should also be a superset of the metadata recorded as part of software citation.
These statements may also be true for software products.
In general, we intend the software citation principles to cover the minimum of what is necessary for software citation for the purpose of software identification.
Other use cases (e.g., provenance, reproducibility) may lead to additional requirements (i.e., enhanced metadata).
%\ssnote{from the data side, an example might be found in the National Climate Assessment (NASA, GCIS) where this is captured in PROV instead?}
%\dmnote{ Mention that the principles are covering what's necessary, not necessarily what is sufficient for a given use case. So perhaps we should discuss things like configurations or bugs? }

\subsection{Software papers} \label{sec:software_papers}

@@

\subsection{Derived software}

The goals of software include the linked ideas of crediting those responsible for software and understanding the dependencies of scientific products on specific software.
In the Importance principle (\ref{principle:importance}), we state that
``software should be cited whenever and wherever a research product (such as a paper or derived software) relies upon it, specifically, as part of the standard reference list for that research product.''
In the case of one code that is derived from another code, citing the derived software may appear to not credit those responsible for the original software, nor recognize its role in the work that used the derived software.
However, this is really analogous to how any research builds on other research, where each research product just cites those products that it directly builds on, not those that it indirectly builds on.
Understanding these chains of knowledge and credit have been part of the history of science field for some time, though more recent work is suggesting more nuanced evaluation of the credit chains~\cite{casrai-credit, transitive_credit_json-ld}.

\subsection{Software peer review}

@@

\subsection{Citations in text}

@@

\subsection{Unique identification}

@@

\subsection{Access to software}

The Accessibility principle (\ref{principle:accessibility}) states that ``software citations should permit and facilitate access to the software itself.''
This does not mean that the software must be freely available.
Rather, the metadata should provide enough information that the software can be accessed.
If the software is free, the metadata will likely provide an identifier that can be resolved to a URL that contains the specific version of the software that is being cited.
For commercial software, the metadata should still provide information on how to access the specific software, but this may be a link to a web site that allows the software be purchased or a company's product number.
As stated in the Persistence principle (\ref{principle:persistence}), we recognize that the software version may no longer be available, but it still should be cited along with information about how it was accessed.

@@

\subsection{Updates to this document}

@@

Jay Jay Billings, Oak Ridge National Laboratory

@@

Scott Edmunds, GigaScience, BGI Hong Kong

>>>>

These software citation principles were originally based on an adaptation of the FORCE11 Data Citation Principles~\cite{data-citation-principles}, and then were modified based on working group discussions and the information from the use cases and the related work.
The adaptations and modifications have been made because software, while similar to data in terms of not traditionally having been cited in publications, is also different than data in that it can be used to express or explain concepts and it is executable.  \katznote{are there any other difference between software and data that should be stated here?} \asnote{...and is likely updated more frequently? (I don't have a reference for this statement)}  Also, while software can be considered a type of data, the converse is not generally true.

@@

With close to 50 working group participants (\ref{app:wg_members}) representing a range of research domains, the working group was tasked to
document existing practices in their respective communities. A total of 47 documents were submitted by working group
participants, with the life sciences, astrophysics, and geosciences being particularly well-represented in the submitted
resources.

@@

Some of the most actionable work has come from the Software Sustainability Institute (SSI) in the form of blog posts written by their community fellows:

@@

Wilson suggests that software authors include a `CITATION' file that documents exactly how the
authors of the software would like to be cited by others~\cite{ssi-citation-files}. While this isn't a formal metadata specification (it's not machine
readable for example) this does offer a solution for authors wishing to give explicit instructions to potential citing authors and as noted in the introduction (\S\ref{sec:intro}), there is evidence that authors follow instructions if they exist ~\cite{10.1371/journal.pone.0136631}.

@@

Most noteable of the community efforts are those of WSSSPE~\cite{wssspe} and SSI~\cite{ssi-workshops} who between them have run a series of workshops aimed at gathering together community memberrs with an interest in improving the status of software in academia. \asnote{Not sure this is a sufficient one-line summary but couldn't think of anything better.} In each of the three years that WSSSPE workshops have run thus far, the participants have produced a report documenting the topics covered. Section 5.8 and Appendix J in \cite{WSSSPE3} has some preliminary work and discussion particularly relevant to this working group.

@@

One approach to increasing the `citability' of software is to encourage the
submission of papers in standard journals describing a piece of research
software. While some journals (e.g., F1000Research) have traditionally accepted
software submissions, the American Astronomical Society (AAS) have recently
announced they will accept software papers in their journals
\cite{aas-sofware-papers}. As the publishers of journals and convenors of domain-specific conferences the authors note the potential importance of professional societies in effecting changes in community practice.

In astronomy \& astrophysics: ASCL \cite{ascl}, AAS journals and the joint
AAS/GitHub workshop \cite{aas-software-index} dedicated to software citation,
indexing and discoverability.

Life sciences/NIH: Software Discovery Index Meeting Report \cite{software-discovery-index} a cross-disciplinary meeting
but with a largely NIH-focus. Suggested minimal information about software (MIAS) standard. Discusses how a common set of metadata
fields are critical for useful indexing - suggested candidates for this list.

Geosciences: Ontosoft \cite{ontosoft} - A Community Software Commons for the Geosciences. Lots of attention given to the metadata
required to describe, discover and execute research software. NSF geosciences workshop around data lifecycle, management and citation \cite{nsf-geo-data}. This report includes many recommendations for data citation.
\LJHnote{ The Seismological Research Letters, Electronic Seismologists publishes papers on software.} %text needs to be cleaned up LJH%

@@

Author guidelines in F1000Research, APA authors, SSI `publish or be damned'
\cite{ssi-publish-or-be-damned} manifesto, workflow for publishing code using
GitHub and Zenodo \cite{github-citable-code-guide}.
\dmnote{There are many other journals publishing software descriptions, some even specialized in software, e.g. \href{http://www.openresearchcomputation.com}{Open Research Computation}.}

\subsection{Existing efforts around metadata standards}

@@

\subsection{Studies of author motivations}

@@

\textbf{What software to cite}

The software citation principles do not define what software should be cited, but rather, how software should be cited.
What software should be cited is the decision of the author(s) of the research work, in the context of community norms and practices, and in most scientific communities, these are currently in flux.
In general, we believe that any software on which a research product directly depends should be cited, if it significantly contributes to the research product.
Again, the specific decision of what is significant must be made by the author(s) of the product.
However, an illustrative example is the use of Microsoft Excel in research.
We suggest that if Excel is used to simply store and plot data, it does not need to be cited, but if it is used for statistical analysis, it should be.  Similarly, general software for conducting library research (e.g., JSTOR Mobile App), writing research papers (e.g., \LaTeX), scientific presentations (e.g., Powerpoint) or communications (e.g., Skype) should not be cited.''
This recommendation matches that of the Purdue Online Writing Lab: ``Do not cite standard office software (e.g. Word, Excel) or programming languages.  Provide references only for specialized software.''~\cite{powl-citing-software}

Note that some software which is or could be captured as part of data provenance may not be cited.
Citation is a record of software that is important to the research outcome, where provenance is a record of all steps (including software) used to generated particular data within the research process.
This implies that for a data research product, provenance data will include all cited software, but not necessarily vice versa.
Similarly, the software metadata that is recorded as part of data provenance should be a superset of the metadata recorded as part of software citation.
And the data recorded for reproducibility should also be a superset of the metadata recorded as part of software citation.
These statements may also be true for software products.
In general, we intend the software citation principles to cover what is minimum of what is necessary for software citation for the purpose of software identification.
Other use cases (e.g., provenance, reproducibility) may lead to additional requirements (i.e., enhanced metadata).
%\ssnote{from the data side, an example might be found in the National Climate Assessment (NASA, GCIS) where this is captured in PROV instead?}
%\dmnote{ Mention that the principles are covering what's necessary, not necessarily what is sufficient for a given use case. So perhaps we should discuss things like configurations or bugs? }

\textbf{Software papers}

@@

\textbf{Derived software}

The goals of software include the linked ideas of crediting those responsible for software and understanding the dependencies of scientific products on specific software.
In the Importance principle (\ref{principle:importance}), we state that
``software should be cited whenever and wherever a research product (such as a paper or derived software) relies upon it, specifically, as part of the standard reference list for that research product.''
In the case of one code that is derived from another code, citing the derived software may appear to not credit those responsible for the original software, nor recognize its role in the work that used the derived software.
However, this is really analogous to how any research builds on other research, where each research product just cites those products that it directly builds on, not those that it indirectly builds on.
Understanding these chains of knowledge and credit have been part of history of science field for some time, though more recent work is suggesting more nuanced evaluation of the credit chains~\cite{casrai-credit, transitive_credit_json-ld}.

\textbf{Software peer review}

@@

\textbf{Citations in text}

@@

\textbf{Unique identification}

@@

\textbf{Access to software}

The Accessibility principle (\ref{principle:accessibility}) states that ``software citations should permit and facilitate access to the software itself.''
This does not mean that the software must be freely available.
Rather, the metadata should provide enough information that the software can be accessed.
If the software is free, the metadata will likely provide an identifier that can be resolved to a URL that contains the specific version of the software that is being cited.
For commercial software, the metadata should still provide information on how to access the specific software, but this may be a link to a web site that allows the software be purchased or a company's product number.
As stated in the Persistence principle (\ref{principle:persistence}), we recognize that the commercial software version may no longer be available, but it still should be cited along with information about how it was accessed.

@@

\textbf{Updates to this document}

@@

@@
